---
title: "Lab 3"
author: "Andrew Steinkruger"
format: html
editor: visual
---

```{r preliminaries, warning = FALSE, message = FALSE}

# Packages

library(tidyverse)
library(gt)
library(spatstat)

# Data

data <- "data/geog565_lab3_data.csv" %>% read_csv

```

**1.1**

```{r 1_1}

# Plots

vis_1_1 = 
  data %>% 
  mutate(group = 
           group %>% 
           factor %>% 
           fct_relevel("commiphora", "termites", "acacia")) %>% 
  ggplot() +
  geom_point(aes(x = x,
                 y = y,
                 color = group)) +
  scale_x_continuous(limits = c(0, 10),
                     breaks = c(0, 10)) +
  scale_y_continuous(limits = c(0, 10),
                     breaks = c(0, 10)) +
  facet_wrap(~ group, ncol = 3) +
  coord_fixed() +
  theme_minimal() +
  theme(legend.position = "none")

vis_1_1

```

-   The *Commiphora* shrubs are clustered over space: most points are close to a small subset of other points and far from all other points.

-   The termite mounds are dispersed over space: most points share near-identical nearest-neighbor distances.

-   The *Acacia senegal* trees are randomly distributed over space: distances between points vary substantially with no clear pattern.

**1.2**

```{r 1_2, warning = FALSE, message = FALSE}

dat_quads = 
  tibble(x = 
           seq(1, 10) %>% 
           ifelse(nchar(.) == 1, 
                  paste0("0", .),
                  .),
         y = 
           seq(1, 10) %>% 
           ifelse(nchar(.) == 1, 
                  paste0("0", .),
                  .),) %>% 
  expand(x, y) %>% 
  mutate(quad = paste0("(", x, ", ", y, ")")) %>% 
  select(quad)

par_N = dat_quads %>% nrow
par_lambda = 150 / 100

dat_1_2 = 
  data %>% 
  mutate(quad_x = 
           x %>% 
           ceiling %>% 
           ifelse(nchar(.) == 1, 
                  paste0("0", .),
                  .),
         quad_y = 
           y %>% 
           ceiling %>% 
           ifelse(nchar(.) == 1, 
                  paste0("0", .),
                  .),
         quad = paste0("(", quad_x, ", ", quad_y, ")")) %>% 
  group_by(group, quad) %>% 
  summarize(count = n()) %>% 
  ungroup %>% 
  arrange(group, quad)

dat_1_2_wrangle = 
  dat_1_2 %>% 
  group_by(group) %>% 
  expand(dat_quads) %>% 
  ungroup %>% 
  left_join(dat_1_2) %>% 
  mutate(count = count %>% replace_na(0)) %>% 
  group_by(group, count) %>% 
  summarize(O = n()) %>% 
  ungroup %>% 
  mutate(P = exp(-par_lambda) * par_lambda ^ count * (1 / factorial(count)),
         E = P * par_N,
         aggregate = 
           c(seq(1, 4), rep(5, 4), # Acacia
             c(1, 2, 3, 3, 4, 4, 5, 6, 6, 6, 6), # Commiphora
             c(1, 1, 2, 3))) %>% # Termites
  group_by(group, aggregate) %>% 
  mutate(O_agg = sum(O),
         E_agg = sum(E)) %>% 
  ungroup %>% 
  mutate(Chi_partial = (O_agg - E_agg) ^ 2 * (1 / E_agg))
  
dat_1_2_chi = 
  dat_1_2_wrangle %>% 
  distinct(group, Chi_partial) %>% # Fragile
  group_by(group) %>% 
  summarize(Chi_sq = sum(Chi_partial),
            degrees = n() - 2) %>% 
  ungroup %>% 
  mutate(value = c(7.815, 9.488, 3.841), # Fragile
         reject = Chi_sq > value,
         group =
           group %>%
           factor %>%
           fct_relevel("commiphora", "termites", "acacia")) %>%
  arrange(group)

tab_1_2_1 = 
  dat_1_2_wrangle %>% 
  filter(group == "commiphora") %>% 
  select(-group) %>% 
  mutate(across(everything(), ~ round(.x, 2))) %>% 
  gt %>% 
  tab_header(title = md("Intermediate Results | *Commiphora* Shrubs")) %>% 
  cols_label(count = html("<em>x</em>"),
             O = html("<em>O<sub>i</sub></em>"),
             P = html("<em>p(x)</em>"),
             E = html("<em>E<sub>i</sub></em>"),
             aggregate = html("<em>j</em>"),
             O_agg = html("<em>O<sub>j</sub></em>"),
             E_agg = html("<em>E<sub>j</sub></em>"),
             Chi_partial = html("Partial &#967;<sup>2</sup>"))

tab_1_2_2 = 
  dat_1_2_wrangle %>% 
  filter(group == "termites") %>% 
  select(-group) %>% 
  mutate(across(everything(), ~ round(.x, 2))) %>% 
  gt %>% 
  tab_header(title = md("Intermediate Results | Termite Mounds")) %>% 
  cols_label(count = html("<em>x</em>"),
             O = html("<em>O<sub>i</sub></em>"),
             P = html("<em>p(x)</em>"),
             E = html("<em>E<sub>i</sub></em>"),
             aggregate = html("<em>j</em>"),
             O_agg = html("<em>O<sub>j</sub></em>"),
             E_agg = html("<em>E<sub>j</sub></em>"),
             Chi_partial = html("Partial &#967;<sup>2</sup>"))

tab_1_2_3 = 
  dat_1_2_wrangle %>% 
  filter(group == "acacia") %>% 
  select(-group) %>% 
  mutate(across(everything(), ~ round(.x, 2))) %>% 
  gt %>% 
  tab_header(title = md("Intermediate Results | *Acacia senegal* Trees")) %>% 
  cols_label(count = html("<em>x</em>"),
             O = html("<em>O<sub>i</sub></em>"),
             P = html("<em>p(x)</em>"),
             E = html("<em>E<sub>i</sub></em>"),
             aggregate = html("<em>j</em>"),
             O_agg = html("<em>O<sub>j</sub></em>"),
             E_agg = html("<em>E<sub>j</sub></em>"),
             Chi_partial = html("Partial &#967;<sup>2</sup>"))

tab_1_2_chi =
  dat_1_2_chi %>% 
  mutate(across(c(Chi_sq, value), ~ round(.x, 2))) %>% 
  gt %>% 
  tab_header(title = html("&#967;<sup>2</sup> Results")) %>% 
  cols_label(group = "Group",
             Chi_sq = html("&#967;<sup>2</sup>"),
             degrees = "DF",
             value = html("Critical Value"),
             reject = "Reject")
  
```

The following tables report statistics by group (e.g. *Commiphora* shrubs) for each quadrat count.

Column "*j*" indexes aggregations of *x* such that each *j* has a sum of *x* equal to or greater than 5.

Columns "*O~j~*", "*E~j~*", and "Partial Chi-Squared" then report results for each *j* (with repetition).

```{r 1_2_print_1}

tab_1_2_1

tab_1_2_2

tab_1_2_3


```

The following table reports chi-squared values and tests of significance for each group.

```{r 1_2_print_2}

tab_1_2_chi

```

**1.3**

For each group, I am testing the null hypothesis of complete randomness by comparing the observed chi-squared to a critical value of the chi-squared distribution with reference to the available degrees of freedom and selected *p*-value. If the observed value is greater than the critical value, then I can reject the null hypothesis that the observed points are distributed by a homogeneous planar Poisson point process and thus that the points are distributed under complete randomness.

Column "Reject" in the preceding table reports the results of these comparisons as logical values.

**1.4**

Please refer to the preceding table for chi-squared values.

-   For *Commiphora* shrubs, I reject the null hypothesis.

-   For termite mounds, I reject the null hypothesis.

-   For *Acacia senegal* trees, I cannot reject the null hypothesis.

Note that these results conform to the visual results in (1.1).

**1.5**

With reference to provided values for variance and lambda:

-   *Commiphora* shrubs are clustered.

-   Termite mounds are dispersed.

-   *Acacia senegal* trees appear to be clustered, but our hypothesis testing suggests this is spurious.

**1.6**

Quadrat analysis requires points that are given, not selected by the researcher, as well as sufficient observations for the chi-squared test. Quadrat analysis also requires the researcher to choose the grain, extent, and position of quadrats. That is, quadrat analysis cannot detect patterns at multiple scales and could suffer from edge effects at the boundaries of each quadrat.

**2.1**

```{r 2_1, warning = FALSE, message = FALSE}

# Get histograms of nearest-neighbor distances.

dat_nn_com =
  ppp(data %>% filter(group == "commiphora") %>% pull(x), 
      data %>% filter(group == "commiphora") %>% pull(y), 
      window = owin(c(0, 10), c(0, 10))) %>% 
  nndist %>% 
  tibble(NN = .,
         Group = "commiphora")

dat_nn_ter = 
  ppp(data %>% filter(group == "termites") %>% pull(x), 
      data %>% filter(group == "termites") %>% pull(y), 
      window = owin(c(0, 10), c(0, 10))) %>% 
  nndist %>% 
  tibble(NN = .,
         Group = "termites")

dat_nn_aca = 
  ppp(data %>% filter(group == "acacia") %>% pull(x), 
      data %>% filter(group == "acacia") %>% pull(y), 
      window = owin(c(0, 10), c(0, 10))) %>% 
  nndist %>% 
  tibble(NN = .,
         Group = "acacia")

dat_nn = 
  dat_nn_com %>% 
  bind_rows(dat_nn_ter) %>% 
  bind_rows(dat_nn_aca)

vis_nn = 
  dat_nn %>% 
  mutate(Group = 
           Group %>% 
           factor %>% 
           fct_relevel("commiphora", "termites", "acacia")) %>% 
  ggplot() +
  geom_histogram(aes(x = NN,
                     fill = Group)) +
  labs(x = "Nearest-Neighbor Distance",
       y = "Count") +
  facet_wrap(~ Group, ncol = 3) +
  theme_minimal() +
  theme(legend.position = "none")

vis_nn
  
```

-   The *Commiphora* shrubs have shorter nearest-neighbor distances centered around 0.25, suggesting clustering.

-   The termite mounds have moderate nearest-neighbor distances in a distribution centered around 0.75 with little variance, suggesting dispersal.

-   The *Acacia senegal* trees have a wide range of nearest-neighbor distances in a distribution centered around 0.50 with greater variance than the other two distributions, suggesting randomness.

**2.2 - 2.6**

In the following table, variables prefixed "Obs\_" are observed values (**2.2**) while "Exp_Mean" is the expected mean (**2.3**). For each group, I am testing the null hypothesis of complete randomness in nearest-neighbor distances (**2.4**). "Z" reports Z-statistics (**2.5**). "Reject" reports the results of hypothesis tests for *p* \< 0.05; for *Commiphora* shrubs and termite mounds, I reject the null hypothesis because each group's Z-statistic outside the range of critical values (-1.96, 1.96), while for *Acacia senegal* trees, I cannot reject the null hypothesis because the group's Z-statistic falls inside the range of critical values (**2.6**).

```{r 2_2}

tab_nn = 
  dat_nn %>% 
  mutate(Group = 
           Group %>% 
           factor %>% 
           fct_relevel("commiphora", "termites", "acacia")) %>% 
  group_by(Group) %>% 
  summarize(Obs_Mean = mean(NN),
            Obs_Var = var(NN),
            Obs_SE = sd(NN) / sqrt(n()),
            Exp_Mean = 0.5 * sqrt(100 / n()),
            Z = (Obs_Mean - Exp_Mean) / Obs_SE) %>% 
  mutate(Reject = abs(Z) > 1.96) %>% 
  mutate(across(2:6, ~ round(.x, 2))) %>% 
  gt

tab_nn

```

**2.7**

-   *Commiphora* shrubs are clustered because their observed mean nearest-neighbor distance is less than the expected mean nearest-neighbor distance and the Z-statistic testing this relationship indicates significance at *p* \< 0.05.

-   Termite mounds are dispersed because their observed mean nearest-neighbor distance is greater than the expected mean nearest-neighbor distance and the Z-statistic testing this relationship indicates significance at *p* \< 0.05.

-   *Acacia senegal* trees appear to be clustered by comparison of the observed and expected means, but our hypothesis testing suggests this is spurious.

**2.8**

Nearest-neighbor analysis requires points that are given, not selected by the researcher, and is subject to bias from edge effects.

**3.1**

```{r 3_1}

par(mfrow = c(1, 3), bty = 'n')

# Commiphora

dat_nn_com %>% 
  pull(NN) %>% 
  sort %>%
  plot(., 
       1:length(.),
       xlab = 'r (distance m)',
       ylab = 'F(r)', main = 'commiphora',
       pch = 21, bg = "blue")

# Termites

dat_nn_ter %>% 
  pull(NN) %>% 
  sort %>%
  plot(., 
       1:length(.),
       xlab = 'r (distance m)',
       ylab = 'F(r)', main = 'termites',
       pch = 21, bg = "blue")

# Acacia

dat_nn_aca %>% 
  pull(NN) %>% 
  sort %>%
  plot(., 
       1:length(.),
       xlab = 'r (distance m)',
       ylab = 'F(r)', main = 'acacia',
       pch = 21, bg = "blue")

par(mfrow = c(1, 1))

```

*r* is a variable radius. *F(r)* is the complete distribution function of all observed nearest-neighbor distances for each *r*.

**3.2**

-   *Commiphora* shrubs appear to be clustered because the corresponding *F(r)* increases steeply for lower values of *r*, indicating that most distances are shorter rather than longer.

-   Termite mounds appear to be dispersed because the corresponding *F(r)* increases steeply for higher values of *r*, indicating that most distances are longer rather than shorter. We can also note relatively little variance in *F(r)* over values of *r*, which corresponds to similar observations in quadrat and NN analysis.

-   *Acacia senegal* trees are not clearly clustered or dispersed because the corresponding *F(r)* is not clearly different than the cumulative distribution function of a normal distribution.

**3.3**

The clustering of *Commiphora* shrubs appears to occur on a scale of around 0.25 m, while the dispersal of termite mounds appears to occur on a scale of around 0.75 m. There is no clear pattern on any relevant scale for *Acacia senegal* trees.

**4.1**

RNN measures the CDF of NN distances, while Ripley's K second-order analysis counts pairs of points (not only nearest neighbors) occurring at different distances.

**4.2**

```{r 4_2, message = FALSE}

# Get plots for Ripley's K.

par(mfrow = c(1, 3), bty = 'n')

ppp(data %>% filter(group == "commiphora") %>% pull(x), 
    data %>% filter(group == "commiphora") %>% pull(y), 
    window = owin(c(0, 10), c(0, 10))) %>% 
  envelope(., 
           fun = Kest, 
           confidence = 0.95,
           correction = "Ripley", 
           var.approx = TRUE) %>% 
  plot(main = 'commiphora')

ppp(data %>% filter(group == "termites") %>% pull(x), 
    data %>% filter(group == "termites") %>% pull(y), 
    window = owin(c(0, 10), c(0, 10))) %>% 
  envelope(., 
           fun = Kest, 
           confidence = 0.95,
           correction = "Ripley", 
           var.approx = TRUE) %>% 
  plot(main = 'termites')

ppp(data %>% filter(group == "acacia") %>% pull(x), 
    data %>% filter(group == "acacia") %>% pull(y), 
    window = owin(c(0, 10), c(0, 10))) %>% 
  envelope(., 
           fun = Kest, 
           confidence = 0.95,
           correction = "Ripley", 
           var.approx = TRUE) %>% 
  plot(main = 'acacia')

par(mfrow = c(1, 1))

```

*r* (or *d* in other notation) is a variable radius or distance as in RNN. *L(r)* counts pairs of points at different distances and transforms the result with the area, sample size, and correction for handling distances over the circle defined by *r* on a two-dimensional plane. This avoids problems with heterogeneity and edge effects inherent to quadrat analysis, NN, and RNN.

**4.3**

The 0.95 confidence interval envelope for the null hypothesis of complete randomness is drawn from a subset of Monte Carlo simulations of the data under complete randomness. That is, the envelope captures the *r*-*L(r)* distributions that are most likely to occur under complete randomness. So, comparison of the observed distributions to the envelope of simulated distribution indicates the likelihood of the observed distribution occurring under complete randomness. The difference of integrals over values of *r* indicates whether an observed distribution is clustered or dispersed relative to complete randomness.

With that, we can note that *Commiphora* shrubs are clustered and (by visual comparison only) that clustering appears to be significant. Termite mounds are dispersed and that dispersal appears to be significant. *Acacia senegal* trees are not clearly clustered or dispersed and it appears that complete randomness holds in their distribution.

**4.4**

The difference in *L(r)* distributions for each plot suggests that *Commiphora* shrubs are clustered at scales from 0.25 to 2.00 units of distance, termites are dispersed at distances of up to 0.75 units of distance, and *Acacia senegal* trees have no interpretable scale of pattern.

**4.5**

Both RNN and Ripley's K reveal patterns across the entire distribution of spatial relationships in a sample, so that both methods support analysis of multiple scales of pattern. Ripley's K has the additional advantages of correcting for heterogeneity and edge effects.

**5.1.a**

Bruce and Tom could use any of the point pattern analysis methods we have examined to investigate the spatial pattern of infections. RNN might be useful in revealing the distances or scales that the infection spreads over.

**5.1.b.**

Bruce and Tom could use Ripley's K to relate their point pattern of interest to another point variable in a bivariate point pattern analysis.

**5.2.a**

Laurie's survey appears to have created 1-dimensional quadrats along transects (unless we add some assumptions about the arrangement of roads or Laurie's imputation of roadside observations to locations further from roads). Since she does not observe the actual points of plant presence/absence, no point pattern analysis method except quadrat analysis could be appropriate, and that would require adaptation of the quadrat methods discussed here to the 1-dimensional transect setting.

**5.2.b**

I do not see an appropriate application of point pattern analysis methods to the association of spatial patterns of exotic plants in a quadrat format with other variables.

**5.3.a**

Shira could use NN and/or RNN analysis of locations of conflicts to identify a univariate spatial pattern of conflicts.

**5.3.b**

Shira could code conflicts by their distance from basin boundaries, from cities, or by features of the nearest basin and/or city, and then develop a Ripley's K second-order analysis on bivariate data.\
