---
title: "Lab 1 | GEOG 565"
author: "Andrew Steinkruger"
format: pdf
editor: visual
---

```{r preliminaries, message = FALSE, warning = FALSE}

library(tidyverse)
library(gt)

dat = "data/geog565_lab1_data.csv" %>% read_csv

vec_1 = dat$precip
vec_2 = dat$streamflow

vec_1_1 = vec_1[1:480]
vec_1_2 = vec_1[481:649]
vec_1_3 = vec_1[650:961]
vec_1_4 = vec_1[962:1465]

vec_2_1 = vec_2[1:480]
vec_2_2 = vec_2[481:649]
vec_2_3 = vec_2[650:961]
vec_2_4 = vec_2[962:1465]

vec_string = c("Dec 1-Dec 20", "Dec 21-Dec 27", "Dec 28-Jan 10", "Jan 11-Jan 31")

```

**1.**

```{r 1}

dat = dat %>% mutate(time = row_number())

dat %>% 
  pivot_longer(cols = c(precip, streamflow),
               values_to = "value",
               names_to = "variable") %>% 
  ggplot() +
  geom_point(aes(x = time,
                 y = value,
                 color = variable),
             alpha = 0.50) +
  labs(x = "Hour (From December 1)", y = "Value", color = "Variable") +
  theme_minimal()

```

**2.**

a.  We can observe temporal autocorrelation in both variables: for each observation, the value is closer to that of the preceding and following observations than another random observation or the mean of all other observations. That is, values for streamflow observations are more similar to those of other observations that are closer rather than further in time, and the same is true for precipitation.

    Borrowing from slides, the distances between streamflow values over time suggest local variances at a distance of 1 time unit vary systematically from global variance pooled at all distances.

b.  Let $Y_t$ be streamflow observed at time $t\in\{1, 1465\}$. Let $n\in\mathbb{Z}$ be a lag in time. Then autocorrelation means $Y_t = F(Y_{t-n})$.

c.  We have that semivariance, lag covariance, and lag correlation measure autocorrelation.

**3.**

a.  We can observe a weak correlation between precipitation and streamflow.

b.  Let $t\in\{1, 1465\}$ be time. Let $X_t$ be precipitation and $Y_t$ be streamflow observed at $t$. Then correlation can be interpreted as $Y_t=F(X_t)$. We might note that the inverse, $X_t=G(Y_t)$ where $G()$ inverts $F()$, is misleading interpretation.

c.  We have that covariance and correlation measure correlation.

**4.**

a.  We can observe strong cross-correlation between precipitation and streamflow, in the sense that changes in streamflow follow changes in precipitation on a lag in time.

b.  Let $t\in\{1, 1465\}$ be time. Let $n\in\mathbb{Z}$ be a lag in time. Let $X_t$ be precipitation and $Y_t$ be streamflow observed at $t$. Then cross-correlation can be interpreted as $Y_t = F(X_{t-n})$. This is equivalent to $Y_{t+n} = F(X_t)$.

c.  We have that cross-covariance and cross-correlation measure cross-correlation.

**5.**

a.  

```{r 5a}

val_5a_precip = vec_1 %>% var %>% round(2)
val_5a_stream = vec_2 %>% var %>% round(2)

```

The variance of precipitation is `{r} val_5a_precip`.

The variance of streamflow is `{r} val_5a_stream`.

b.  

```{r 5b}

val_5b_precip = vec_1 %>% sd %>% round(2)
val_5b_stream = vec_2 %>% sd %>% round(2)

```

The standard deviation of precipitation is `{r} val_5b_precip`.

The standard deviation of streamflow is `{r} val_5b_stream`.

c.  

```{r 5c}

val_5c_precip = vec_1 %>% mean %>% round(2)
val_5c_stream = vec_2 %>% mean %>% round(2) 

```

The mean of precipitation is `{r} val_5c_precip`.

The mean of streamflow is `{r} val_5c_stream`.

d.  

```{r 5d}

val_5d_precip = ((vec_1 %>% sd) / (vec_1 %>% mean)) %>% round(2)
val_5d_stream = ((vec_2 %>% sd) / (vec_2 %>% mean)) %>% round(2)

```

The coefficient of variation for precipitation is `{r} val_5d_precip`.

The coefficient of variation for streamflow is `{r} val_5d_stream`.

e.  Variance is calculated from mean deviations, so it is centered by definition.

f.  Variance takes the squared units of the underlying distribution (e.g. squared unit volume per squared unit time for streamflow), so it is non-normalized by definition.

g.  The coefficient of variation is normalized, so we can compare the two variables. CV for precipitation is greater than that of streamflow, so precipitation has more variation. Someone watching a creek through a rainstorm might intuit the same result.

**6.**

```{r 6}

dat_6 = 
  tibble('Time Period' = vec_string,
         Correlation = 
           c(cor(vec_1_1, vec_2_1) %>% round(2),
             cor(vec_1_2, vec_2_2) %>% round(2),
             cor(vec_1_3, vec_2_3) %>% round(2),
             cor(vec_1_4, vec_2_4) %>% round(2)))

dat_6 %>% gt %>% tab_options(latex.use_longtable = TRUE)

```

In the first three periods, precipitation and streamflow exhibit weak positive correlations. In the fourth and final period, precipitation and streamflow exhibit a weak negative correlation. Eyeballing the plot in (1), these correlations follow from the lagged relationship between the two variables. In the first three periods, precipitation events lead to a small instantaneous increase in streamflow, while in the fourth period that relationship falls apart because (almost) no further precipitation events occur and streamflow is falling on a longer time scale (seasonal) than our unit of analysis (hourly).

Values differ among time periods; as above, this is because precipitation almost stops after hour 1,000.

{{< pagebreak >}}

**7.**

a.  

```{r 7a}

ccf(vec_1_1, vec_2_1, lag.max = 24, main = vec_string[1], ylab = "Cross-Correlation")
ccf(vec_1_2, vec_2_2, lag.max = 24, main = vec_string[2], ylab = "Cross-Correlation")
ccf(vec_1_3, vec_2_3, lag.max = 24, main = vec_string[3], ylab = "Cross-Correlation")
ccf(vec_1_4, vec_2_4, lag.max = 24, main = vec_string[4], ylab = "Cross-Correlation")

```

{{< pagebreak >}}

b.  

```{r 7b}

dat_7 = 
  tibble('Lag (Hours)' = c(1, 3, 6, 12, 24),
         'Dec 1-Dec 20' = 
           ccf(vec_1_1, 
               vec_2_1, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(24, 22, 19, 13, 1)] %>% 
           round(2),
         'Dec 21-Dec 27' = 
           ccf(vec_1_2, 
               vec_2_2, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(24, 22, 19, 13, 1)] %>% 
           round(2),
         'Dec 28-Jan 10' = 
           ccf(vec_1_3, 
               vec_2_3, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(24, 22, 19, 13, 1)] %>% 
           round(2),
         'Jan 11-31' = 
           ccf(vec_1_4, 
               vec_2_4, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(24, 22, 19, 13, 1)] %>% 
           round(2))

dat_7 %>% 
  gt %>% 
  tab_spanner(label = "Cross-Correlation", 
              columns = 
                c('Dec 1-Dec 20',
                  'Dec 21-Dec 27',
                  'Dec 28-Jan 10',
                  'Jan 11-31')) %>% 
  tab_options(latex.use_longtable = TRUE)

```

c.  The plots suggest that, for the first three periods, the cross-correlation of precipitation and streamflow increases up to a maximum at a lag between 0 and 11 hours, then decreases as the lag becomes a lead and increases. This matches the intuitive notion of a waterway picking up in flow as its watershed collects precipitation over time; the period of 0-11 hours reflects the spatial scale (and slope, among other characteristics) of this particular waterway and watershed. In the fourth period, with little additional precipitation and a larger-scale trend of decreasing streamflow following a seasonal maximum, the cross-correlation is uninformative.

**8.**

a.  

```{r 8a}

fun_acf = function(vec, label){
  
  plot(acf(vec, lag.max = 24, plot = FALSE),
       type = "l",
       col = "blue",
       ylab = "Autocorrelation",
       main = label) 
  
}

fun_acf(vec_1_1, label = "Precipitation (Dec 1-Dec 20)")
fun_acf(vec_1_2, label = "Precipitation (Dec 21-Dec 27)")
fun_acf(vec_1_3, label = "Precipitation (Dec 28-Jan 10)")
fun_acf(vec_1_4, label = "Precipitation (Jan 11-Jan 31)")

fun_acf(vec_2_1, label = "Streamflow (Dec 1-Dec 20)")
fun_acf(vec_2_2, label = "Streamflow (Dec 21-Dec 27)")
fun_acf(vec_2_3, label = "Streamflow (Dec 28-Jan 10)")
fun_acf(vec_2_4, label = "Streamflow (Jan 11-Jan 31)")

```

b.  

```{r 8b1}

dat_8b1 = 
  tibble('Lag (Hours)' = c(1, 3, 6, 12, 24),
         'Dec 1-Dec 20' = 
           acf(vec_1_1, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(2, 4, 7, 13, 25)] %>% 
           round(2),
         'Dec 21-Dec 27' = 
           acf(vec_1_2, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(2, 4, 7, 13, 25)] %>% 
           round(2),
         'Dec 28-Jan 10' = 
           acf(vec_1_3, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(2, 4, 7, 13, 25)] %>% 
           round(2),
         'Jan 11-31' = 
           acf(vec_1_4, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(2, 4, 7, 13, 25)] %>% 
           round(2))

dat_8b1 %>% 
  gt %>% 
  tab_spanner(label = "Precipitation Autocorrelation", 
              columns = 
                c('Dec 1-Dec 20',
                  'Dec 21-Dec 27',
                  'Dec 28-Jan 10',
                  'Jan 11-31')) %>% 
  tab_options(latex.use_longtable = TRUE)

```

```{r 8b2}

dat_8b2 = 
  tibble('Lag (Hours)' = c(1, 3, 6, 12, 24),
         'Dec 1-Dec 20' = 
           acf(vec_2_1, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(2, 4, 7, 13, 25)] %>% 
           round(2),
         'Dec 21-Dec 27' = 
           acf(vec_2_2, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(2, 4, 7, 13, 25)] %>% 
           round(2),
         'Dec 28-Jan 10' = 
           acf(vec_2_3, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(2, 4, 7, 13, 25)] %>% 
           round(2),
         'Jan 11-31' = 
           acf(vec_2_4, 
               lag.max = 24, 
               plot = FALSE)
         [[1]][c(2, 4, 7, 13, 25)] %>% 
           round(2))

dat_8b2 %>% 
  gt %>% 
  tab_spanner(label = "Streamflow Autocorrelation", 
              columns = 
                c('Dec 1-Dec 20',
                  'Dec 21-Dec 27',
                  'Dec 28-Jan 10',
                  'Jan 11-31')) %>% 
  tab_options(latex.use_longtable = TRUE)

```

{{< pagebreak >}}

**9.**

a.  With reference to the first table under (8b), autocorrelation in precipitation decreases from (almost) 1 to (almost) 0 over lags from 1 to 24 hours. This suggests precipitation events in the study area and period tend to fall off over a 24 hour period. These values differ over periods because precipitation events differ over the given temporal bins; we can note that the precipitation events we observe in the figure under (1) are heterogeneous in their intensity and duration.

b.  With reference to the second table under (8b), autocorrelation in streamflow decreases from (almost) 1 to (almost) 0 following precipitation events and remains between 0.50 and 1.00 in periods with little precipitation, like January 11-31. This suggests past streamflow has a positive relationship with present streamflow that weakens over time. These values differ over periods because precipitation events differ, as in (9a); here, physical attributes of waterways yield stronger autocorrelations over the temporal bins.

c.  With reference to the table under (6), correlations between precipitation and streamflow are positive for the first three periods and negative in the last period. This captures only the instantaneous relationship between precipitation and streamflow, as well as some noise from lagged effects of precipitation on streamflow. Correlation differs among periods because precipitation events of different intensity have instantaneous effects of different intensity, and noise from lagged effects is non-random.

d.  With reference to the table under (7b), cross-correlations between precipitation and streamflow increase to a maximum at a lag of 1-12 hours and decrease thereafter, except in the fourth period, where cross-correlations are relatively small and negative. This captures the lagged effects I raise in (9c): precipitation has some non-zero instantaneous effect on streamflow, then an increasing lagged effect as rainfall is collected in the observed waterway. These effects differ over periods because precipitation events vary in intensity and, plausibly, in their spatial distributions. \
    \
    With the evidence from (a)-(d), we can conclude that if precipitation is a driver and streamflow is a resopnse, then their relationship in this data demonstrates a modulated response, a lagged response, and hysteresis. We might observe additional behaviors with a different time series or different temporal bins.

**10.**

The three statistics of interest differ in their information on univariate/bivariate and instantaneous/lagged relationships. Autocorrelation represents univariate relationships over time lags. Correlation represents instantaneous bivariate relationships. Cross-correlation represents bivariate relationships over time lags. For the time series at hand, each statistic is useful in parsing different features of the relationship between precipitation and streamflow, as observed in the different conclusions from (9a) through (9d).
